\vspace{-2.5em} Finding degeneracy in fluorescence orientation microscopes \vspace{-1em}

### 2018-04-02 notes/2018-03-31-degeneracy-search/report/report.tex
SO3 -> O3

We have a set of $N$ equations that we need to solve simultaneously for
$\so{} \in \mathbb{S}^2$ and $\mb{R} \in \mathbb{SO}(3)$. As far as I can tell
there is no analytic way to solve this set of equations in the general case. I
tried expanding onto spherical and rotational harmonics, but that didn't help
because there's no easy way to relate a solution in the frequency domain to a
solution in the normal domain. Instead, we can find as many solutions as we can
``by hand'' then check our results numerically. I've used optimization packages
the can optimize on $\mathbb{S}^2\times \mathbb{SO}(3)$---my rough plan is to
seed an optimization procedure at many place on a grid in
$\mathbb{S}^2\times \mathbb{SO}(3)$ and check (not prove!) that we have all of
the solutions with those results. Other ideas welcome!

For now let's solve a simple case ``by hand''. If we have a single point
detector along the $z$ axis then our angular point spread function is given by
\begin{align}
  h^{(\mh{z})}(\so{}) = 1 - (\so{}\cdot\mh{z})^2. \label{eq:zpsf}
\end{align}
This is the first time I'm expressing an angular point spread function in vector
notation, but it's easy to check that this reduces to the familiar $h^{(\mh{z})}(\so{}) = \sin^2\vartheta$ if you choose spherical coordinates so that $\so{} = \cos\varphi\sin\vartheta\mh{x} + \sin\varphi\sin\vartheta\mh{y} + \cos\vartheta\mh{z}$. Plugging Eq. \ref{eq:zpsf} into Eq. \ref{eq:tosolve} gives
\begin{align}
  -(\so{}\cdot\mh{z})^2 + (\mb{R}^{-1}\so{}\cdot\mh{z})^2 = 0.
\end{align}
I can find the following solutions to this equation for all $\so{} \in \mathbb{S}^2$
\begin{align}
  \mb{R}_0 =
  \begin{bmatrix}
    \cos\theta& -\sin\theta & 0\\
    \sin\theta& \cos\theta & 0\\
    0&0&1\\
  \end{bmatrix} \text{for}\ 0 \leq \theta < 2\pi,
\end{align}
which is an arbitrary rotation about the $z$ axis;
\begin{align}
  \mb{R}_1 = \mb{I} + 2 \left[\mb{k}\right]_{\times}^2
\end{align}
where
\begin{align}
  \mb{k} = \frac{\so{}\times[(\so{}\cdot\mh{z})\mh{z}]}{|\so{}\times[(\so{}\cdot\mh{z})\mh{z}]|}, \qquad  \left[\mb{k}\right]_{\times} =
  \begin{bmatrix}
    0& -k_z& k_y\\
    k_z& 0 & -k_x\\
    -k_y& k_x& 0 \\
  \end{bmatrix},
\end{align}
which is a rotation that maps $\so{}$ to $-\so{}$; and
\begin{align}
  \mb{R}_2 = \mb{I} + 2(\so{}\cdot\mh{z})\sqrt{1 - (\so{}\cdot{z})^2}\left[\mb{k}\right]_{\times} + 2(\so{}\cdot\mh{z})^2\left[\mb{k}\right]_{\times}^2
\end{align}
which is a rotation that maps every $\so{}$ above the $x-y$ plane to an $\so{}$
below the $x-y$ plane and vice versa. Notice that
\begin{align}
  \mb{R}_3 =
  \begin{bmatrix}
    -1&0&0\\
    0&-1&0\\
    0&0&-1\\    
  \end{bmatrix}
\end{align}
is not a solution because $\mb{R}_3$ is not a member of $\mathbb{SO}(3)$---$\mb{R}_3$ is a member of the larger set $\mathbb{O}(3)$. This distinction is important
so that we don't double count solutions.

@@

>>>>

We have a set of $N$ equations that we need to solve simultaneously for
$\so{} \in \mathbb{S}^2$ and $\mb{R} \in \mathbb{O}(3)$. Notice that we allow
for improper rotations ($\text{det}\mb{R} = -1$) by allowing $\mb{R} \in \mathbb{O}(3)$
instead of restricting to $\mb{R} \in \mathbb{SO}(3)$.

As far as I can tell there is no analytic way to solve this set of equations in
the general case. I tried expanding onto spherical and rotational harmonics, but
that didn't help because there's no easy way to relate a solution in the
frequency domain to a solution in the normal domain. Instead, we can find as
many solutions as we can ``by hand'' then check our results numerically. I've
used optimization packages the can optimize on
$\mathbb{S}^2\times \mathbb{O}(3)$---my rough plan is to seed an optimization
procedure at many place on a grid in $\mathbb{S}^2\times \mathbb{O}(3)$ and
check (not prove!) that we have all of the solutions with those results. Other
ideas welcome!

For now let's solve a simple case ``by hand''. If we have a single point
detector along the $z$ axis then our angular point spread function is given by
\begin{align}
  h^{(\mh{z})}(\so{}) = 1 - (\so{}\cdot\mh{z})^2. \label{eq:zpsf}
\end{align}
This is the first time I'm expressing an angular point spread function in vector
notation, but it's easy to check that this reduces to the familiar $h^{(\mh{z})}(\so{}) = \sin^2\vartheta$ if you choose spherical coordinates so that $\so{} = \cos\varphi\sin\vartheta\mh{x} + \sin\varphi\sin\vartheta\mh{y} + \cos\vartheta\mh{z}$. Plugging Eq. \ref{eq:zpsf} into Eq. \ref{eq:tosolve} gives
\begin{align}
  -(\so{}\cdot\mh{z})^2 + (\mb{R}^{-1}\so{}\cdot\mh{z})^2 = 0.
\end{align}
I can find the following solutions to this equation for all $\so{} \in \mathbb{S}^2$
\begin{align}
  \mb{R}_0 =
  \begin{bmatrix}
    \cos\theta& -\sin\theta & 0\\
    \sin\theta& \cos\theta & 0\\
    0&0&1\\
  \end{bmatrix} \text{for}\ 0 \leq \theta < 2\pi,
\end{align}
which is an arbitrary rotation about the $z$ axis;
\begin{align}
  \mb{R}_1 =
  \begin{bmatrix}
    1&0&0\\
    0&1&0\\
    0&0&-1\\    
  \end{bmatrix},
\end{align}
which is a mirroring operation about the transverse plane; and
\begin{align}
  \mb{R}_2 =
  \begin{bmatrix}
    -1&0&0\\
    0&-1&0\\
    0&0&-1\\    
  \end{bmatrix},
\end{align}
which is an inversion operation.

@@

There are more solutions to the problem---for example a proper rotation that
maps $\so{}$ to $-\so{}$ instead of an improper rotation---but these solutions
give the same mappings as the solutions above, so it's okay to ignore them.
Alternatively, we could restrict ourselves to finding solutions in
$\mathbb{SO}(3)$ to find unique solutions, but this makes the solutions much
more difficult to identify.
